High Resolution Identification of Protein-DNA Binding Events and Quality Control for ChIP-exo data

### 2015-11-14 rmarkdown/chip_exo_paper.Rnw
Added segment to paper about enrichment and library complexity and fixed figures to include labels

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/for_paper/EColi_strand_cross_corr.pdf}
  \caption{Strand cross-correlation curves for $\sigma^{70}$
    samples.}
  \label{figsup:scc}
\end{figure}

@@

@@

In ChIP-exo experiments and input samples is often unavailable,
therefore to determine the quality of a ChIP-exo experiment is
necesary to address the balance between the enrichment and the library
complexity of the experiment. To diagnose this balance we considered
the Average Read Coverage (ARC) and the Unique Read Coverage Rate (URCR) which are defined as:

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/for_paper/Sig70_enrichment.pdf}
  \includegraphics[width = .8\textwidth,page =3]{../figs/for_paper/Local_SCC_indicator_by_strata.pdf}
  \caption{A) For all $\sigma^{70}$ ChIP-exo experiments we are
    comparing the ARC against the URCR.}
  \label{fig:enrich}
\end{figure}

@@

We next evaluated the ``peak pair'' assumption of Rhee and Pugh
\cite{exo1}, MACE \cite{MACE}, CexoR, \cite{cexor} and peakzilla
\cite{peakzilla}, i.e. a peak of reads in the forward strand is
usually paired with a peak of reads in the reverse strand that is
located in the other site of the binding site.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width = .6\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  % \includegraphics[width = .9\textwidth ,page = 3]{../figs/for_paper/MA_plot_fwd_bwd_by_seq.pdf}
  \caption{ A) ChIP signal in ChIP-exo was linearly related to that in
    PET ChIP-Seq data in the region with high ChIP tag counts. In
    contrast, there wer clear differences in their background
    distribution, where several ChIP-exo background regions were
    empty.}
  \label{fig:comp}
\end{figure}

@@

\nocite{exo_gb}

>>>>

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/for_paper/EColi_strand_cross_corr.pdf}
  \caption{Strand cross-correlation curves for $\sigma^{70}$
    samples. The ``phantom peak'' and the summit that corresponds to
    the read and fragment length respectively are confounded, due to
    the exonuclease enzyme digestion.}
  \label{figsup:scc}
\end{figure}

@@

In ChIP-exo experiments, the background is often digested by the
exonuclease enzyme, therefore to determine the sample's quality is
necesary to address the balance between the enrichment and library
complexity. To diagnose this, we considered the Average Read Coverage
(ARC) and the Unique Read Coverage Rate (URCR) which are defined as:

Figure \ref{fig:enrich}A shows hexbin plots illustrating the
interaction between these two quantities. Notice how there are two
strong arms in each panel: The first one corresponds to regions with
low $\mbox{ARC}$ values and $\mbox{URCR}$ values that can vary accross
the whole $[0,1]$ interval, while the second one shows a decreasing
trend in $\mbox{URCR}$ as $\mbox{ARC}$ increases. When an experiment
shows a higher degree of enrichment, then the separation of this two
arms is more noticeable, since the second arm corresponds to possibly
enriched regions (\ref{fig:enrich2}A and \ref{fig:enrich2}B).

<<strata_param , include =FALSE,echo = FALSE, eval = TRUE>>=

  nstrata <- 100

Figure \ref{fig:enrich}B shows boxplots of the $\mbox{local-NSC}$
stratified by the number of reads mapped exactly to one position for
all $\sigma^{70}$ ChIP-exo experiments; for each stratum, we sampled
\Sexpr{nstrata} regions that whenever possible (in the opposite case,
all the regions in the category were considered). There is an
increasing trend as the number of unique position increases, which
means that this is an effective indicator of library complexity since
regions where the reads are mapped to more positions are usually more
complex.

%% perhaps do two other slots about how this works... a good experiment vs a bad experiment
%% of the same region

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/for_paper/Sig70_enrichment.pdf}
  \includegraphics[width = .8\textwidth,page =3]{../figs/for_paper/Local_SCC_indicator_by_strata.pdf}
  \caption{A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$ and
    B) boxplots of the $\mbox{local-NSC}$ stratified by nr. of reads
    mapped to only one position in the regions for the ChIP-exo
    $\sigma^{70}$ experiments divided by condition and biological
    replicate}
  \label{fig:enrich}
\end{figure}

@@

@@

We next evaluated the ``peak pair'' assumption of Rhee and Pugh
\cite{exo1}, MACE \cite{mace}, CexoR, \cite{cexor} and peakzilla
\cite{peakzilla}, i.e. a peak of reads in the forward strand is
usually paired with a peak of reads in the reverse strand that is
located in the other site of the binding site.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width = .6\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  \caption{ A) ChIP signal in ChIP-exo was linearly related to that in
    PET ChIP-Seq data in the region with high ChIP tag counts. In
    contrast, there wer clear differences in their background
    distribution, where several ChIP-exo background regions were
    empty.}
  \label{fig:comp}
\end{figure}

@@

\nocite{exo_gb}
\nocite{maplot1}
\nocite{maplot2}

\section*{Supplement}
\label{sec:supp}

\subsection*{Additional Enrichment plots for $\sigma^{70}$}
\label{sec:enrichsup}

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth,page = 3]{../figs/for_paper/Sig70_enrichment.pdf}
  \includegraphics[width = .8\textwidth,page = 4]{../figs/for_paper/Sig70_enrichment.pdf}
  \caption{Hexbin plots of $\mbox{ARC}$ vs $\mbox{URCR}$ for each
    region after partitioning the genome. In A and B the regions with
    reads mapped to at most 10 and 30 positions respectively where not
    considered.}
  \label{fig:enrich2}
\end{figure}

\subsection{Local-NSC}
\label{sec:localnsc}

The local strand cross-correlaiton is defined as:

\begin{align}
  \mbox{local-NSC} &= \frac{max_{x_\delta} f(x_\delta)}{\sigma} \nonumber \\
 y_\delta &= f(x_\delta) + \sigma \epsilon_\delta \nonumber
\end{align}

where $f$ is calculated by fitting a non-parametric regression model
over the strand cross-correlation calculated using exclusively the
fragments that overlap a pre-defined genomic region, $y_\delta$ is the
correlation of the forward and backward coverages when one is shifted
$x_\delta$ bp towards their 3' end.

### 2015-11-14 rmarkdown/chip_exo_paper.tex
Added segment to paper about enrichment and library complexity and fixed figures to include labels

>>>>

In ChIP-exo experiments, the background is often digested by the
exonuclease enzyme, therefore to determine the sample's quality is
necesary to address the balance between the enrichment and library
complexity. To diagnose this, we considered the Average Read Coverage
(ARC) and the Unique Read Coverage Rate (URCR) which are defined as:

Figure \ref{fig:enrich}A shows hexbin plots illustrating the
interaction between these two quantities. Notice how there are two
strong arms in each panel: The first one corresponds to regions with
low $\mbox{ARC}$ values and $\mbox{URCR}$ values that can vary accross
the whole $[0,1]$ interval, while the second one shows a decreasing
trend in $\mbox{URCR}$ as $\mbox{ARC}$ increases. When an experiment
shows a higher degree of enrichment, then the separation of this two
arms is more noticeable, since the second arm corresponds to possibly
enriched regions (\ref{fig:enrich2}A and \ref{fig:enrich2}B).

Figure \ref{fig:enrich}B shows boxplots of the $\mbox{local-NSC}$
stratified by the number of reads mapped exactly to one position for
all $\sigma^{70}$ ChIP-exo experiments; for each stratum, we sampled
100 regions that whenever possible (in the opposite case,
all the regions in the category were considered). There is an
increasing trend as the number of unique position increases, which
means that this is an effective indicator of library complexity since
regions where the reads are mapped to more positions are usually more
complex.

%% perhaps do two other slots about how this works... a good experiment vs a bad experiment
%% of the same region

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/for_paper/Sig70_enrichment.pdf}
  \includegraphics[width = .8\textwidth,page =3]{../figs/for_paper/Local_SCC_indicator_by_strata.pdf}
  \caption{A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$ and
    B) boxplots of the $\mbox{local-NSC}$ stratified by nr. of reads
    mapped to only one position in the regions for the ChIP-exo
    $\sigma^{70}$ experiments divided by condition and biological
    replicate}
  \label{fig:enrich}
\end{figure}
